---
title: "Full stack data science in R"
author: "Michał Burdukiewicz"
date: ""
output: 
  beamer_presentation:
    slide_level: 2
    toc: true
    includes:
      in_header: preamble.tex
---

## About me

 - bioinformatician (Warsaw University of Technology, Brandenburg University of Technology Cottbus-Senftenberg, .prot),
 - founder of the Why R? Foundation and Wrocław R User Group,
 - facilitator of McKinsey Tech Meetup.

## Advanced R

Elective subject: "Programowanie w R dla zaawansowanych".

# Full stack data science

## Full stack data science

\begin{figure} 
\includegraphics[width=0.85\textwidth]{static_figure/overview.png}
\end{figure}


## Full stack data science in R

\begin{figure} 
\includegraphics[width=0.85\textwidth]{static_figure/overview-r.png}
\end{figure}


# Data acquisition and processing

## Data acquisition

Import data to **R** in a *data.frame* (or similar) format.

```{r,echo=FALSE}
knitr::kable(data.frame(Type = c("Tabular", "Relational databases", "Graph databases (e.g., neo4j)"),
                        Package = c("**readr**, **xlsx**, **data.table**::*fread*", "**RPostgreSQL**, **mongolite**", "None")),
             booktabs = TRUE)
```

## Big data processing

```{r,echo=FALSE,message=FALSE,warning=FALSE,fig.height=4.3}
library(ggplot2)
library(reshape2)
library(dplyr)
dat <- read.csv2("big.csv", header = FALSE)[,1:3]
colnames(dat) <- c("Functions", "Memory", "Overhead")

present_theme <- theme_bw(base_size = 16) +
  theme(legend.position = "bottom", 
        plot.background = element_rect(fill = NA, color = NA), 
        legend.background= element_rect(fill = NA))

p <- melt(dat, id.vars = "Functions") %>% 
  mutate(value = gsub(pattern = "^ ", "", value),
         value =  as.numeric(sapply(strsplit(as.character(value), " "), first)),
         value = ifelse(value == 660, value/1000, value),
         Functions = factor(Functions, levels = rev(unique(Functions))),
         variable = factor(variable, levels = rev(levels(variable)))) %>%
  arrange(desc(variable)) %>% 
  ggplot(aes(x = Functions, y = value)) +
  geom_col() +
  scale_x_discrete("Functions/packages") +
  scale_y_continuous("RAM used [GB]") +
  #scale_fill_manual("", values = c("#998ec3", "#f1a340")) +
  coord_flip() +
  present_theme

p
```

Comparative performance of matter for linear regression and calculation of the first two principal components on simulated datasets of 1.2 GB. 

\tiny Source: Bemis KA (2019). matter: A framework for rapid prototyping with binary data on disk. R package version 1.10.0, https://github.com/kuwisdelu/matter.


## Big data processing

```{r,echo=FALSE,message=FALSE,warning=FALSE,fig.height=4.3}
p + geom_col(aes(fill = variable)) +
  scale_fill_manual("", values = c("#998ec3", "#f1a340"))
```

Memory overhead is the maximum memory used during the execution minus the memory in use upon completion.

\tiny Source: Bemis KA (2019). matter: A framework for rapid prototyping with binary data on disk. R package version 1.10.0, https://github.com/kuwisdelu/matter.

# Model development

## mlr

**mlr**: a standardized interface to machine learning in **R** (by Bernd Bischl and Michael Lang).

Alternatives: **caret**, **parsnip**.

## Why mlr?

 - amazing documentation,
 - separation of learner and task,
 - a wide array of tasks: *Classification*, *Regression*, *Cost-sensitive*, Survival, Clustering, Multilabel, Imbalanced data, Functional data, Spatial data.
 - 71 performance measures.
 - \huge{MBO}.
 
## mlr tasks

\tiny
```{r,echo=FALSE,message=FALSE}
library(mlr)
```

```{r}
iris[1L:2, ]
iris.task
```

## mlr learners

\tiny
```{r,warning=FALSE}
listLearners()
```

## mlr learners

\tiny
```{r,warning=FALSE}
listLearners(obj = iris.task)
```

## mlr learners

\tiny
```{r,}
iris.task <- makeClassifTask(data = iris, target = "Species")

lrn_rng <- makeLearner("classif.ranger")
# ranger: the fastest random forest implementation in R
# Marvin Wright, the author of ranger, is a keynote speaker of Why R? 2019

cv <- makeResampleDesc("CV", iters = 3)

resample(learner = lrn_rng, task = iris.task, resampling = cv)
```


## Bayesian optimization

$$\max_{x \in \mathbb{R}}f(x)$$

The *posterior* probability of a model (or theory, or hypothesis) $M$ given evidence (or data, or observations) $E$ is proportional to the likelihood of $E$ given $M$ multiplied by the *prior* probability of $M$:

$$P(M|E) \propto P(E|M)|P(M) $$

## Bayesian optimization

Find local extrema of arbitrary functions.

## Bayesian optimization

$y = \sin(x) \times x^2$

Local minimum: -3.9453.

```{r,echo=FALSE,fig.height=4.6}
data.frame(x = seq(-3.8, 2, 0.01)) %>% 
  mutate(y = sin(x)*x^2) %>% 
  ggplot(aes(x = x, y = y)) +
  geom_line() +
  present_theme
```


## Bayesian optimization

```{r,warning=FALSE,message=FALSE,echo=FALSE,include=TRUE,results="hide" }
library(mlrMBO)
set.seed(1590)
fn2 = makeSingleObjectiveFunction(
  name = "Sample Function",
  fn = function(x) sin(x)*x^2,
  par.set = makeNumericParamSet("x", len = 1L, lower = -3.8L, upper = 2L),
  global.opt.params = list(x = 0)
)

ctrl = makeMBOControl(propose.points = 1)
ctrl = setMBOControlTermination(ctrl, iters = 6L)
ctrl = setMBOControlInfill(ctrl, crit = makeMBOInfillCritEI(),
                           opt = "focussearch", opt.focussearch.points = 500L)

run <- exampleRun(fn2, control = ctrl)

# przybli¿ana funkcja + wszystkie badane punkty
df <- as.data.frame(run$mbo.res$opt.path) # wszystkie wytypowane punkty
g1 <- ggplot(data=run$evals, aes(x=x,y=y)) + geom_line() + geom_point(data=data.frame(x=df$x,y=df$y)) + 
  geom_point(data=data.frame(x=run$mbo.res$x,y=run$mbo.res$y), color="red") +
  geom_text(data=data.frame(x=run$mbo.res$x,y=run$mbo.res$y), label = round(run$mbo.res$y, 2), color="red", vjust = -1) +
  ggtitle("MBO") +
  present_theme


# Random Search

library(randomsearch)
runRandomSearch <- randomsearch(fn2, max.evals = 10L)
df2 <- as.data.frame(runRandomSearch)
opt2 <- getOptPathEl(runRandomSearch, index = getOptPathBestIndex(runRandomSearch))
g2 <- ggplot(data=run$evals, aes(x=x,y=y)) + geom_line() + geom_point(data=data.frame(x=df2$x,y=df2$y)) + 
  geom_point(data=data.frame(x=opt2$x,y=opt2$y), color="red") +
  geom_text(data=data.frame(x=opt2$x,y=opt2$y), label = round(opt2$y, 2), color="red", vjust = -1) +
  ggtitle("Random search") +
  present_theme


# Grid Search

library(paramtest)
fn2.b <- function(iter, x) return(sin(x)*x^2);
runGridSearch <- grid_search(fn2.b, params = list(x=seq(-3.8,2)))
df3 <- data.frame(x=runGridSearch$tests$x, y=unlist(runGridSearch$results))
opt3 <- data.frame(x=runGridSearch$tests$x[which.min(unlist(runGridSearch$results))],y=min(unlist(runGridSearch$results)))
g3 <- ggplot(data=run$evals, aes(x=x,y=y)) + geom_line() + geom_point(data=data.frame(x=df3$x,y=df3$y)) + 
  geom_point(data=data.frame(x=opt3$x,y=opt3$y), color="red") +
  geom_text(data=data.frame(x=opt3$x,y=opt3$y), label = round(opt3$y, 2), color="red", vjust = -1) +
  ggtitle("Grid search") +
  present_theme


library(gridExtra)
grid.arrange(g3, g2, g1, ncol = 1)

```

## Baysian optimization

\begin{figure} 
\includegraphics[width=0.85\textwidth]{static_figure/fig1.png}
\end{figure}

Source: Bernd Bischl

# Model deployment

## Shiny

\begin{figure} 
\includegraphics[width=0.95\textwidth]{static_figure/shiny.png}
\end{figure}

Limitations:

\begin{enumerate}
\item stability decreases with the number of concurrent users,
\item speed.
\end{enumerate}

## ShinyProxy and Kubernetes

**ShinyProxy**: docker-based containers for Shiny apps.

ShinyProxy containers can be themselves run in a container and deployed clustered container managers (such as Kubernetes or Swarm).

# Project management

## drake

\texttt{make} for \textbf{R} projects.

*GNU Make is a tool which controls the generation of executables and other non-source files of a program from the program's source files.*

Source: GNU Make web page.

## drake

\small
```{r,eval=FALSE,echo=TRUE}
plan <- drake_plan(
  raw_data = read_excel(file_in("raw_data.xlsx")),
  data = raw_data %>%
    mutate(Species = fct_inorder(Species)),
  hist = create_plot(data),
  fit = lm(Sepal.Width ~ Petal.Width + Species, data),
  report = rmarkdown::render(
    knitr_in("report.Rmd"),
    output_file = file_out("report.html"),
    quiet = TRUE
  )
)
```

## drake

\begin{figure} 
\includegraphics[width=0.85\textwidth]{static_figure/drake1.png}
\end{figure}

## drake

Parallelization:
\small
```{r,eval=FALSE,echo=TRUE}
# multicore scaling
make(plan, jobs = 12)

# or enable supercomputing clusters with 
# the future package 
drake_hpc_template_file("torque_batchtools.tmpl") 
library(future.batchtools)
future::plan(batchtools_torque, 
             template = "torque_batchtools.tmpl", 
             workers = 8000)
make(plan, parallelism = "future_lapply")
```

## Why R? 2019

\begin{figure} 
\includegraphics[width=0.75\textwidth]{static_figure/europa.jpg}
\end{figure}
